iSEE: Interactive SummarizedExperiment Explorer

### 2018-03-29 manuscript/main.tex
Added abstract, tightened up some of the main text.

\abstract{\textbf{Motivation:} Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text
Text Text Text Text Text.\\
\textbf{Results:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text\\
\textbf{Availability:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text  Text\\
\textbf{Contact:} \href{aaron.lun@cruk.cam.ac.uk}{aaron.lun@cruk.cam.ac.uk}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\subsection{Integration with Bioconductor packages} 
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
This object stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively. 
Per-feature or per-sample variables can also be easily added to the object in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results. 
The flexibility of the SummarizedExperiment class is the driving factor behind its wide deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, ChIP-seq \citep{lun2016csaw}, methylation \citep{aryee2014minfi}, Hi-C \citep{lun2016infrastructure} and mass cytometry (PMID: 28504682, lukas). 
Developers can easily derive subclasses for specialized use cases, such as the SingleCellExperiment class that enables easy storage of sample-wise reduced dimensionality embeddings in single-cell `-omics data analyses. 
By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization to complement existing analysis pipelines for a variety of data modalities. 
In particular, we leverage off the state-of-the-art statistical and computational methods already available in the R/Bioconductor framework, allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).

\subsection{Visualizing multiple data aspects simultaneously}
iSEE uses a customisable multi-panel layout that simultaneously displays one or more panels of different types, where each panel type visualizes a different aspect of the data. 
This includes metadata for each feature or sample, experimental observations across samples for each feature, and reduced dimensionality embeddings for each sample. 
For most panels, each sample or feature is represented as a point in a plot with the specified variables on the x/y-axes.
Depending on whether these variables are categorical or numeric, panels will automatically switch between scatter plots, violin plots or ``rectangle plots'' (where each combination of categorical levels is represented by a rectangle with area proportional to the number of samples/features with that combination). 
Each point can also be coloured according to data aspects such as sample-level metadata or feature-level observations. 
Other panel types include heatmaps, to visualize experimental observations for multiple features; and row statistics tables, to facilitate searching across features and their metadata.
All panels can be added, removed, reordered or resized.

@@

\subsection{Customizing the user experience}
iSEE also offers a host of other customizable features to improve data exploration. 
Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each individual  metadata variable and assay type consistently across all panels in the application. 
Row statistics tables can be augmented with dynamic annotation based on the selected row, providing links to online resources such as Ensembl or Entrez annotation. 
For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization. 
After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.

\subsection{Reproducing the plotting code}
iSEE memorizes the exact R code that was used to generate every plot. 
This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts. 
All plots are generated using the ggplot2 framework and can be easily customized.

>>>>

\abstract{\textbf{Summary:} Data exploration is criticial to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
However, most existing tools for interactive visualization are limited to specific assays or analyses.
Here, we present the iSEE (Interactive SummarizedExperiment Explorer) package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
iSEE is directly compatible with many existing R/Bioconductor frameworks for analyzing high-throughput biological data,
and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics datasets.\\
\textbf{Availability:} iSEE is publicly available as an R package from the open-source Bioconductor project (https://bioconductor.org/packages/iSEE).\\
\textbf{Contact:} \href{aaron.lun@cruk.cam.ac.uk}{aaron.lun@cruk.cam.ac.uk}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\subsection{Integration with Bioconductor packages} 
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
This object stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively. 
Per-feature or per-sample variables can also be easily added to the object in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results. 
The flexibility of the SummarizedExperiment class is the driving factor behind its wide deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C \citep{lun2016infrastructure}.
Developers can easily derive subclasses for specialized use cases, such as the SingleCellExperiment class that stores sample-wise reduced dimensionality embeddings in single-cell `-omics data analyses. 
By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization to complement existing analysis pipelines for a variety of data modalities. 
%In particular, we leverage off the state-of-the-art statistical and computational methods already available in the R/Bioconductor framework, allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).

\subsection{Visualizing multiple data aspects simultaneously}
iSEE uses a customisable multi-panel layout that simultaneously displays one or more panels of various types, where each panel type visualizes a different aspect of the data. 
This includes metadata for each feature or sample, experimental observations across samples for each feature, and reduced dimensionality embeddings for each sample. 
For most panels, each sample or feature is represented as a point in a plot with the specified variables on the x/y-axes.
Depending on whether these variables are categorical or numeric, panels will automatically switch between scatter plots, violin plots or ``rectangle plots'' (where each combination of categorical levels is represented by a rectangle with area proportional to the number of samples/features with that combination). 
Each point can also be coloured according to data aspects such as sample-level metadata or feature-level observations. 
Other panel types include heatmaps, to visualize experimental observations for multiple features; and row statistics tables, to facilitate searching across features and their metadata.
All panels can be removed, reordered or resized by the user, and new panels of any type can be added to the interface.

@@

\subsection{Customizable and flexibility}
iSEE also offers a host of other customizable features to improve data exploration. 
Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each individual  metadata variable and assay type consistently across all panels in the application. 
Row statistics tables can be augmented with dynamic annotation based on the selected row, providing links to online resources such as Ensembl or Entrez. 
For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization. 
After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.
iSEE also memorizes the exact R code that was used to generate every plot. 
This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts.
